Computational design of RNA-based oscillatory circuits

### 2015-09-14 MScWriteup.tex
further rewrites

 In contrast to the results of Fig. \ref{InitialResults}, we see relatively tightly estimated parameters, with no estimated parameter values near the initial bounding box. Further, this simplified model achieves error values as low as the more complex model. In fact, Fig \ref{13_9_f_simplified} shows that the simplified algorithm is more consistently able to find the minimal function value identified in Fig. \ref{InitialResults_13_9_f}, possibly because the low dimension of the new model makes parameter space easier to search.

@@

In this report, we have presented a system of ODE's to model a recently proposed synthetic RNA regulatory circuit \cite{Rodrigo2012}. We have attempted to estimate the model's parameters using existing time series data, with a least squares minimisation approach, similar to that found in recent synthetic biology literature \cite{Hu2015}. Where this approach has failed to give estimates, we have discussed why, and suggested that a simpler model may provide an equally good description of the data with fewer parameters than the initial model.

@@

These results suggest that, if estimates of the parameters contained in \eqref{eq:s} - \eqref{eq:c} are desired, fluorescence time series data alone will not be sufficient. One option, though perhaps experimentally infeasible, would be to directly observe other components of the state vector - $s$, $m$, $s:m$ and $c$. This would improve parameter estimates by giving direct observations of the `fast' timescale. On the other hand, if a good description of the fluorescence data is what is really of interest, further work may consist of using the simplified model, but improving the modelling of the translation step.  \cite{Hu2015}, for example, uses a more detailed model than our single step description of translation. 
We also note that $\mu$ and $\Theta$ are both parameters which can be easily estimated independent of our datasets - it may be of interest to compare those values to the ones estimated here.

Further work may also consist of improving our methodology. Using a least squares minimisation approach has several problems, the first of which is local minima. Using the CMA-ES will give better results than a local minimisation algorithm, but one can never be sure that global optimum has been found - for example, \cite{Algorithms2003} compares several algorithms and finds, in the test case considered, none locate the true global optimum. A related problem is that of working with point estimates of parameters rather than distributions. The CMA-ES will only ever find a single optimum, but in the case where the error landscape is very flat -  for example a flat basin pocketed by many local minima - the exact minimum it finds paints a misleading picture, because it gives no idea of the uncertainty in the answer given. To surmount this problem, we may rerun the algorithm many times, from many starting locations, and interpret the resulting spread of parameter values found as an uncertainty \cite{Hu2015}, as we have done in this report. However this methodology is rather ad hoc.

@@

\begin{itemize}
\item systematic mode comparison ? or subset of paramter estimation if we really want to keep this model.
\end{itemize}

@@

       \begin{subfigure}[c]{0.49\textwidth}
        \centering
    \includegraphics[scale = 0.35, clip = true, trim = 0 0 0 0]{13_9_f_hist}
        \caption{Error values found from 200 initial parameter estimates, \textit{13\_9 dataset}. }
        \label{InitialResults_13_9_f}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.35, clip = true, trim = 0 0 0 0]{14_7_f_hist}
        \caption{Error values found from 200 initial parameter estimates, \textit{14\_7 dataset}}
        \label{InitialResults_14_7_f}
    \end{subfigure}
    \caption{Inter-parameter correlation values, and error function values, for the \textit{13\_9} and \textit{14\_7} datasets.}     
\label{InitialResults_additional}
\end{figure}

@@

\section{Initial parameter Estimates, both datasets}
\label{Bothdatasetfitting}
\setcounter{figure}{0}

\begin{figure}[H]
    \begin{subfigure}[c]{\textwidth}
    \centering
        \includegraphics[scale = 0.3]{13_9_14_7_with_13_9_bestplot}
        \caption{Model prediction from parameter set which gives lowest error when trained on the combination of both datasets, plotted against the  \textit{13\_9} dataset. }
    \end{subfigure}

    \begin{subfigure}[c]{\textwidth}
    \centering
        \includegraphics[scale = 0.3]{13_9_14_7_with_14_7_bestplot}
        \caption{Model prediction from parameter set which gives lowest error when trained on the combination of both datasets, plotted against the  \textit{14\_7} dataset.}
    \end{subfigure}
    \caption{}
\label{hist_f}
\end{figure}

@@

\begin{figure}[H]
   \centering
        \includegraphics[scale = 0.4]{13_9_14_7_hist}
        \caption{Histogram of estimated parameter values, found from 100 runs of the CMA-ES algorithm. Fitted to both the  \textit{13\_9} and \textit{14\_7} datasets.}
        \label{}
\end{figure}

@@

 \begin{figure}[h!]
% ensure that we have normalsize text
\tiny
% Store the current equation number.
\setcounter{MYtempeqncnt}{\value{equation}}
% Set the equation number to one less than the one
% desired for the first equation here.
% The value here will have to changed if equations
% are added or removed prior to the place these
% equations are referenced in the main text.
\setcounter{equation}{11}
\begin{align}
\label{eq:m_fixed}  
m &= \frac{1}{{2 k_{\text{on}} \left(\mu +\delta _m\right)
   \left(k_{\text{hyb}}+\delta _m\right)}}
   \Bigg(\sqrt{2 (\text{am}+\text{as}) k_{\text{on}} \left(\mu +\delta _m\right) \left(\mu +\delta _s\right) \left(k_{\text{hyb}}+\delta _m\right)
   \left(k_{\text{hyb}}+k_{\text{off}}+\delta _m\right)+(\text{am}-\text{as})^2 k_{\text{on}}^2 \left(k_{\text{hyb}}+\delta _m\right){}^2+\left(\mu +\delta _m\right){}^2
   \left(\mu +\delta _s\right){}^2 \left(k_{\text{hyb}}+k_{\text{off}}+\delta _m\right){}^2} \\ \nonumber  
 &- (\text{am}-\text{as}) k_{\text{on}} \left(k_{\text{hyb}}+\delta
   _m\right)+\left(\mu +\delta _m\right) \left(\mu +\delta _s\right) \left(k_{\text{hyb}}+k_{\text{off}}+\delta _m\right)\Bigg)
\end{align}
\begin{align}
\label{eq:c_fixed}
c &= \frac{k_{\text{hyb}}}{2 k_{\text{on}} \left(\mu +\delta
   _m\right) \left(k_{\text{hyb}}+\delta _m\right){}^2}
   \Bigg( \sqrt{2 (\text{am}+\text{as}) k_{\text{on}} \left(\mu +\delta _m\right) \left(\mu +\delta _s\right) \left(k_{\text{hyb}}+\delta _m\right)
   \left(k_{\text{hyb}}+k_{\text{off}}+\delta _m\right)+(\text{am}-\text{as})^2 k_{\text{on}}^2 \left(k_{\text{hyb}}+\delta _m\right){}^2+\left(\mu +\delta _m\right){}^2
   \left(\mu +\delta _s\right){}^2 \left(k_{\text{hyb}}+k_{\text{off}}+\delta _m\right){}^2} \\ \nonumber
   &+(\text{am}+\text{as}) k_{\text{on}} \left(k_{\text{hyb}}+\delta
   _m\right)+\left(\mu +\delta _m\right) \left(\mu +\delta _s\right) \left(k_{\text{hyb}}+k_{\text{off}}+\delta _m\right) \Bigg)
\end{align}
% Restore the current equation number.
\setcounter{equation}{\value{MYtempeqncnt}}
% IEEE uses as a separator
\hrulefill
% The spacer can be tweaked to stop underfull vboxes.
\vspace*{4pt}
\end{figure}

>>>>

 In contrast to the results of Fig. \ref{InitialResults}, we see relatively tightly estimated parameters, with no estimated parameter values near the initial bounding box. Further, this simplified model achieves error values as low as the more complex model. Fig \ref{13_9_f_simplified} shows that the simplified algorithm is more consistently able to find the minimal function value identified in Fig. \ref{InitialResults_13_9_f}, possibly because the low dimension of the new model makes parameter space easier to search.

@@

In this report, we have presented a system of ODE's to model a recently proposed synthetic RNA regulatory circuit \cite{Rodrigo2012}, and attempted to estimate the model's parameters using existing time series data, with a least squares minimisation approach similar to that found in recent synthetic biology literature \cite{Hu2015}. We have found many of the parameters contained in the full model to be inestimable, and suggested reasons  for why this might be. We have further suggested a simplified model, which provides an equally good description of the data with fewer parameters than the initial model.

@@

These results suggest that, if estimates of the parameters contained in \eqref{eq:s} - \eqref{eq:c} are desired, fluorescence time series data alone may not be sufficient. One option, though perhaps experimentally infeasible, would be to directly observe other components of the state vector - $s$, $m$, $s:m$ and $c$. This would improve parameter estimates by giving direct observations of the `fast' timescale. Another option would be further investigation of the methods presented in \cite{Mclean2012,Yao2003,Mclean2012-2}. Given a complex chemical model based on physical principles, and limited data, these methods rank parameters in order of estimability, and can determine if a subset of parameters might be estimable.

On the other hand, if a good description of the fluorescence data is what is really of interest, further work may consist of reformulating the model, perhaps beginning with the simplified model in \eqref{eq:p_simplified} - \eqref{eq:z_simplified}, but improving the modelling of the translation step.  \cite{Hu2015}, for example, uses a more detailed model than our single step description of translation. When constructing this revised model, a systematic approach to model comparison, 
\todo[inline]{talk about AIC, bayesian mode comparison here? https://sciencehouse.files.wordpress.com/2013/07/gordonconf13.pdf}
We also note that $\mu$ and $\Theta$ are both parameters which can be easily estimated independent of our datasets - it may be of interest to compare those values to the ones estimated here.

Further work may also consist of improving our methodology for parameter estimation. Using a least squares minimisation approach has several problems, the first of which is local minima. Using the CMA-ES will give better results than a local minimisation algorithm, but one can never be sure that global optimum has been found - for example, \cite{Algorithms2003} compares several algorithms and finds, in the test case considered, none locate the true global optimum. A related problem is that of working with point estimates of parameters rather than distributions. The CMA-ES will only ever find a single optimum, but in the case where the error landscape is very flat -  for example a flat basin pocketed by many local minima - the exact minimum it finds paints a misleading picture, because it gives no idea of the uncertainty in the answer given. To surmount this problem, we may rerun the algorithm many times, from many starting locations, and interpret the resulting spread of parameter values found as an uncertainty \cite{Hu2015}, as we have done in this report. However this methodology is rather ad hoc.

@@

\begin{itemize}
\item systematic mode comparison ? or subset of paramter estimation if we really want to keep this model.
\item literature bounds?
\end{itemize}

@@

       \begin{subfigure}[c]{0.49\textwidth}
        \centering
    \includegraphics[scale = 0.35, clip = true, trim = 0 0 0 0]{13_9_f_hist}
        \caption{Error values found from 200 initial parameter estimates, \textit{13\_9 dataset}. }
        \label{InitialResults_13_9_f}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.35, clip = true, trim = 0 0 0 0]{14_7_f_hist}
        \caption{Error values found from 200 initial parameter estimates, \textit{14\_7 dataset}}
        \label{InitialResults_14_7_f}
    \end{subfigure}
    \caption{Inter-parameter correlation values and error function values, found from 200 initial parameter estimates, for the \textit{13\_9} and \textit{14\_7} datasets.}     
\label{InitialResults_additional}
\end{figure}

@@

\section{Initial parameter Estimates, fit using both datasets}
\label{Bothdatasetfitting}
\setcounter{figure}{0}

\begin{figure}[H]
    \begin{subfigure}[c]{\textwidth}
    \centering
        \includegraphics[scale = 0.3]{13_9_14_7_with_13_9_bestplot}
        \caption{Model prediction from the parameter set which gives lowest error when trained on the combination of both datasets, plotted against the  \textit{13\_9} dataset. }
    \end{subfigure}

    \begin{subfigure}[c]{\textwidth}
    \centering
        \includegraphics[scale = 0.3]{13_9_14_7_with_14_7_bestplot}
        \caption{Model prediction from the parameter set which gives lowest error when trained on the combination of both datasets, plotted against the  \textit{14\_7} dataset.}
    \end{subfigure}
    \caption{}
\label{hist_f}
\end{figure}

@@

\begin{figure}[H]
   \centering
        \includegraphics[scale = 0.4]{13_9_14_7_hist}
        \caption{Histograms of estimated parameter values, found from 100 runs of the CMA-ES algorithm. Fitted to both the  \textit{13\_9} and \textit{14\_7} datasets.}
        \label{}
\end{figure}

@@

 \begin{figure}[h!]
% ensure that we have normalsize text
\tiny
% Store the current equation number.
\setcounter{MYtempeqncnt}{\value{equation}}
% Set the equation number to one less than the one
% desired for the first equation here.
% The value here will have to changed if equations
% are added or removed prior to the place these
% equations are referenced in the main text.
\setcounter{equation}{14}
\begin{align}
\label{eq:m_fixed}  
m &= \frac{1}{{2 k_{\text{on}} \left(\mu +\delta _m\right)
   \left(k_{\text{hyb}}+\delta _m\right)}}
   \Bigg(\sqrt{2 (\text{am}+\text{as}) k_{\text{on}} \left(\mu +\delta _m\right) \left(\mu +\delta _s\right) \left(k_{\text{hyb}}+\delta _m\right)
   \left(k_{\text{hyb}}+k_{\text{off}}+\delta _m\right)+(\text{am}-\text{as})^2 k_{\text{on}}^2 \left(k_{\text{hyb}}+\delta _m\right){}^2+\left(\mu +\delta _m\right){}^2
   \left(\mu +\delta _s\right){}^2 \left(k_{\text{hyb}}+k_{\text{off}}+\delta _m\right){}^2} \\ \nonumber  
 &- (\text{am}-\text{as}) k_{\text{on}} \left(k_{\text{hyb}}+\delta
   _m\right)+\left(\mu +\delta _m\right) \left(\mu +\delta _s\right) \left(k_{\text{hyb}}+k_{\text{off}}+\delta _m\right)\Bigg)
\end{align}
\begin{align}
\label{eq:c_fixed}
c &= \frac{k_{\text{hyb}}}{2 k_{\text{on}} \left(\mu +\delta
   _m\right) \left(k_{\text{hyb}}+\delta _m\right){}^2}
   \Bigg( \sqrt{2 (\text{am}+\text{as}) k_{\text{on}} \left(\mu +\delta _m\right) \left(\mu +\delta _s\right) \left(k_{\text{hyb}}+\delta _m\right)
   \left(k_{\text{hyb}}+k_{\text{off}}+\delta _m\right)+(\text{am}-\text{as})^2 k_{\text{on}}^2 \left(k_{\text{hyb}}+\delta _m\right){}^2+\left(\mu +\delta _m\right){}^2
   \left(\mu +\delta _s\right){}^2 \left(k_{\text{hyb}}+k_{\text{off}}+\delta _m\right){}^2} \\ \nonumber
   &+(\text{am}+\text{as}) k_{\text{on}} \left(k_{\text{hyb}}+\delta
   _m\right)+\left(\mu +\delta _m\right) \left(\mu +\delta _s\right) \left(k_{\text{hyb}}+k_{\text{off}}+\delta _m\right) \Bigg)
\end{align}
% Restore the current equation number.
\setcounter{equation}{\value{MYtempeqncnt}}
% IEEE uses as a separator
\hrulefill
% The spacer can be tweaked to stop underfull vboxes.
\vspace*{4pt}
\end{figure}
